**fabryka**

### 2025-07-30 README.Rmd
minor changes in article text after second round of review in PCI archaeology

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15880861.svg)](https://doi.org/10.5281/zenodo.15880861)

@@

fabryka is an open access R *shiny* user-friendly web application dedicated to the fabric analysis of archaeological or geological assemblage. fabryka uses classical analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical tests on orientations) for analysing the fabrics of objects, an spatial method and new ways of spatial and statistical exploration. A descriptive article is available at <https://zenodo.org/records/15880861>.

@@

* clone the fabryka repository from GitHub: <https://github.com/marchaeologist/fabryka> or download the code in "Code" and "Download ZIP" at the same address. This will give you the latest version of the source code.

@@

A tutorial is included in the application (*i.e.* 'Tutorial" panel) and can also be found in a dedicated article: <https://zenodo.org/records/15880861>.
'fabryka' includes an example file you can download from the application ('Upload data' subpanel).

@@

THOMAS, M. (2025). Fabryka: A web application for easily analysing & exploring the fabric of archaeological assemblages. In PCI archaeology (Version 1). Zenodo. https://doi.org/10.5281/zenodo.15880861

>>>>

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.16601423.svg)](https://doi.org/10.5281/zenodo.16601423)

@@

fabryka is an open access R *shiny* user-friendly web application dedicated to the fabric analysis of archaeological or geological assemblage. fabryka uses classical analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical tests on orientations) for analysing the fabrics of objects, an spatial method and new ways of spatial and statistical exploration. A descriptive article is available at <https://zenodo.org/records/16601423>.

@@

* clone the fabryka repository from GitHub: <https://github.com/marchaeologist/fabryka> or download the code in "Code" and "Download ZIP" at the same address. Unzip the folder. This will give you the latest version of the source code.

@@

A tutorial is included in the application (*i.e.* 'Tutorial" panel) and can also be found in a dedicated article: <https://zenodo.org/records/16601423>.
'fabryka' includes an example file you can download from the application ('Upload data' subpanel).

@@

THOMAS, M. (2025). Fabryka: A web application for easily analysing & exploring the fabric of archaeological assemblages. In PCI archaeology (Version 1). Zenodo. https://doi.org/10.5281/zenodo.16601423

### 2025-07-30 fabryka_pci/article.Rmd
minor changes in article text after second round of review in PCI archaeology

---
title: '**fabryka**'
subtitle: A web application for easily analysing & exploring the fabric of archaeological assemblages
output:
  pdf_document: default
header-includes:
- \usepackage{amsmath}
- \usepackage{lineno}
- \linenumbers
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{fabryka}
- \fancyhead[R]{R web application for fabric analysis}
- \fancyfoot[C]{\thepage}
- \usepackage{float}
- \floatplacement{figure}{H}
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

@@

This article presents fabryka 1.1, an open access R shiny web
application dedicated to the fabric analysis of archaeological or
geological assemblages. The fabric is a statistic made of the
orientation and dip of elongated particles and archaeological objects.
It provides information about the nature of the geological processes
involved in the formation of the deposits and their impact on the
preservation of the archaeological assemblages. These statistical
analysis are not easily accessible to non-specialists and are
particularly time-consuming, sometimes requiring the use of several
software, some of which are chargeable. fabryka uses classical
analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical
tests on orientations) for analysing the fabrics of objects and includes
a spatial method. New statistical methods and new ways of exploration
are proposed in the application and this present article.\smallbreak

@@

Fabric is made of the orientation and dip of elongated objects
[@bertran1997; @bertran2002; @lenoble2004]. It is a statistic (*i.e.*,
it needs as much measurements as possible) that enables the
identification of sedimentary processes involved in the formation of the
deposits [@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be considered as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, analysis of the
fabric of sedimentary deposits depends mainly on archaeological material
(bones and lithics). This makes it possible to identify the sedimentary
processes to which they were subjected and to discuss their degree of
spatial disorganization.

@@

Two main families of statistical method are commonly used to
characterize the fabric of a given assemblage [@bertran2002;
@mcpherron2018; @de2021new].

The first methods focus on the orientation
of the measurements only and aim to test the null hypothesis of a random
orientation from a uniform distribution through several statistical
tests [@krumbein1939; @curray1956; @rao1967large].

The second statistical method take into account object orientation and
dip. Three normalized eigenvalues (E1, E2 and E3) are calculated from
the cosine formed by the axis of the artifact with the three axes of
space [@watson1966]. @woodcock1977 and @benn1994 indices are used to
describe the fabric shape. Several authors *e.g.*, [@bertran1997;
@bertran2002; @lenoble2004] built models to compare the fabric of
archaeological assemblages to natural sedimentary processes and human
processes (*i.e.*, fabric of knapping spots). The null hypothesis of an
absence of perturbation is a fabric shape compatible with that of
experimental knapping spots. These methods are described in the
Mathematics section of this article.

Several software allow data processing, some of which are not freely
accessible, do not allow spatial exploration of the data or do not offer
interfaces to facilitate the task of non-programmer users
[@mcpherron2018].

The following sections explain all the selected approaches and our
proposals for processing the fabrics thanks to fabryka, from taking
measurements in the field, through the statistical methods and graphical
modes of representation used in the application (fig.
\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/presentation_applications.png}
  \caption{fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

@@

During excavation, the orientation and dip of elongated remains (*i.e.* a > 2 * b and with a > 2 cm) is
measured on a representative set of artefacts or on the whole
assemblage. The orientation (= bearing) of the lowest point is measured
in the range [0; 360] with North as the reference (*i.e.* orientation =
0°). The dip (= plunge) is measured in the range [0; 90] with horizontal
as a reference. The orientation [0; 360] can be converted into axial
data measured in the range [0; 180] by subtracting 180° from the
measurements in the range ]180; 360].

Traditionally, the orientation of the objects lowest point is measured
using a compass and the dip is measured using an inclinometer. The aim
is to measure the orientation and dip of the physical axis of elongation
of objects ignoring its shape. To do this, the measuring device must be
placed parallel to the general axis of elongation of the object.
Measuring the fabric from its imprint or visible surface is biased by
its shape, making impossible to assess its true dip. Since the
theodolite was introduced to excavations in the 1990s, the fabrics is
sometimes measured by recording the coordinates of the ends of objects [@mcpherron2005artifact].
In the latter case, the rules of trigonometry are applied to convert the
pair of coordinates into orientations and dips.

@@

A sample corresponds to all the orientation and dip measurements taken
within a Lithostratigraphic Unit (LU) or a pedostratigraphic unit. These
units seem to be the most coherent entity which needs to be explored
because they are defined according to geological criterion, which a
priori guarantee a certain coherence in terms of the formation dynamics
we are seeking to approach. In fabryka, analysis are constrained within
samples unless the user decides otherwise in the spatial exploration
panel. \bigbreak

@@

1.  The first considers the orientation only. The calculation of the
    Vector Magnitude $L$ (expressed as a percentage) by Curray's
    [-@curray1956] formula tests the hypothesis of a unimodal
    distribution of orientations. If the $p$ value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the Vector Magnitude $L$ is as follows:

@@

Certain geological processes can have a different impact on objects depending on their
physical properties (*e.g.* silex versus bones in a water flow as in the
example files provided where bones have a strong preferential
orientation and are much more 'linear' in the Benn diagram [@thomas2019; @texier2020]. The natural processes models that you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find
the reference model data (with corresponding literature references)
which is used in the application in the 'Model reference data' subpanel.
In the rose diagram, axial data is considered (and not orientations) as
the Rayleigh test is perform on axial data [@curray1956] in the summary
table. You can download all the figures by selecting a file format
(.pdf, .jpeg and .png) and pressing the 'download' button (fig.
\ref{fig:figure_classical_method}E and F).

@@

-   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source code: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
-   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
-   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15909666> \smallbreak
-   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.15880861> \bigbreak

@@

2. Clone the fabryka repository from GitHub:
<https://github.com/marchaeologist/fabryka> or download the code in
"Code" and "Download ZIP" at the same address. This will give you the
latest version of the source code.

@@

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
Todisco on issues related to site formation processes. The reference
data used to build the models of the classical method were collected by
P. Bertran throughout his career. P. Bertran provided us with all these
data in order to feed the application. The development of the
application also benefited from the help of Rémi Lemoy and François and
Victor Baleux on mathematical issues. A first version of the application
was tested by Emmanuel Discamps and Aurélien Royer. This project is
supported by the "Service Valorisarion de la Recherche" of the
University Toulouse Jean Jaurès. Finally, the constructive comments of
Pascal Bertran, recommender of this article, and, of Frédéric
Santos, Nicolas Frerebeau and Alfonso Benito-Calvo, reviewers of this
article, considerably improved the manuscript and the web application. I
would like to thank all these reaseachers for their advice. \bigbreak

>>>>

---
title: '**fabryka**'
subtitle: A web application for easily analysing & exploring the fabric of archaeological
  assemblages
output:
  word_document: default
  pdf_document:
    keep_tex: true
header-includes:
- \usepackage{amsmath}
- \usepackage{lineno}
- \linenumbers
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{fabryka}
- \fancyhead[R]{R web application for fabric analysis}
- \fancyfoot[C]{\thepage}
- \usepackage{float}
- \floatplacement{figure}{H}
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

@@

This article presents fabryka 1.1, an open access R shiny web
application dedicated to the fabric analysis of archaeological or
geological assemblages. The fabric is a statistic made of the
orientation and dip of elongated particles and archaeological objects.
It provides information about the nature of the geological processes
involved in the formation of the deposits and their impact on the
preservation of the archaeological assemblages. This statistical
analysis is not easily accessible to non-specialists and is
particularly time-consuming, sometimes requiring the use of several
software, some of which are chargeable. fabryka uses classical
analytical procedures (Benn, rose, Schmidt and Woodcock diagrams and statistical
tests on orientations) for analysing the fabrics of objects and includes
a spatial method. New statistical methods and new ways of exploration
are proposed in the application and in this present article.\smallbreak

@@

Fabric is made of the orientation and dip of elongated objects
[@bertran1997; @bertran2002; @lenoble2004]. It is a statistic (*i.e.*,
it needs as much measurements as possible) that enables the
identification of sedimentary processes involved in the formation of the
deposits [@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be considered as sedimentary
particles and are affected by natural processes [@schiffer1983]. As
geological particles are often not elongated enough, analysis of the
fabric of sedimentary deposits often depends mainly on archaeological material
(bones and lithics). This makes it possible to identify the sedimentary
processes to which they were subjected and to discuss their degree of
spatial reorganization. When possible, distinguishing sedimentary particles from archaeological objects in the fabric analysis could allow to test whether these two populations were affected by different processes (*e.g.*, the fabric of an undisturbed occupation covered by deposits from a minor flood should reveal a distinct fabric from that of the alluvial pebbles on which it rests).

@@

Two main families of statistical method are commonly used to
characterize the fabric of a given assemblage [@bertran2002;
@mcpherron2018; @de2021new]. \newline
The first methods focus on the orientation only and aim to test the null hypothesis of a random
orientation of objects through several statistical
tests [@krumbein1939; @curray1956; @rao1967large].\newline
The second statistical method takes into account object orientation and
dip. Three normalized eigenvalues (E1, E2 and E3) are calculated from
the cosine formed by the axis of the artifact with the three axes of
space [@watson1966]. @woodcock1977 and @benn1994 indices are used to
describe the fabric shape. Several authors *e.g.*, [@bertran1997;
@bertran2002; @lenoble2004] built models to compare the fabric of
archaeological assemblages to natural sedimentary processes and human
processes (*i.e.*, fabric of knapping spots). The null hypothesis of an
absence of perturbation is a fabric shape compatible with that of
experimental knapping spots. These methods are described in the
Mathematics section of this article.

Several software allow data processing, some of which are not freely
accessible, do not allow spatial exploration of the data or do not offer
interfaces to facilitate the task of non-programmer users
[@mcpherron2018]. The following sections explain all the selected approaches and our
proposals for processing the fabrics thanks to fabryka, from taking
measurements in the field, through the statistical methods and graphical
modes of representation used in the application (fig.
\ref{fig:figure_presentation}). \bigbreak

\begin{figure} [h]
\centering
  \includegraphics[width=15cm]{figure/presentation_applications.png}
  \caption{fabryka creation scheme and organisation of the article}
\label{fig:figure_presentation}
\end{figure}

@@

During excavation, the orientation and dip of elongated remains (*i.e.* a > 2 * b and with a > 2 cm) is
measured on a representative set of artefacts or on the whole
assemblage. The orientation (= bearing) of the lowest point is measured
in the range [0; 360[ with North as the reference (*i.e.* orientation =
0°). The dip (= plunge) is measured in the range [0; 90] with horizontal
as a reference. The orientation [0; 360[ can be converted into axial
data measured in the range [0; 180[ by subtracting 180° from the
measurements in the range [180; 360].

Traditionally, the azimuth of the principal axis of an object is measured using a compass considering that the point with the lowest altitude indicates the orientation of the vector. The dip is measured using an inclinometer. The aim is to measure the orientation and dip of the elongaton axis of
of objects ignoring its shape. To do this, the measuring device must be
placed parallel to the elongation axis of the object.
Measuring the fabric from its imprint or visible surface is biased by
its shape, making impossible to assess its true dip. Since the
theodolite was introduced to excavations in the 1990s, the fabrics is
sometimes measured by recording the coordinates of the ends of objects [@mcpherron2005artifact].
In the latter case, the rules of trigonometry are applied to convert the
pair of coordinates into orientations and dips.

@@

A sample corresponds to all the orientation and dip measurements taken
within a Lithostratigraphic Unit (LU) or a pedostratigraphic unit. These
units seem to be the most coherent entity which needs to be explored
because they are defined according to geological criteria, which a
priori guarantee a certain coherence in terms of the formation dynamics
we are seeking to approach. In fabryka, analysis are constrained within
samples unless the user decides otherwise in the spatial exploration
panel. \bigbreak

@@

1.  The first considers the orientation only. The calculation of the
    Vector Magnitude $L$ (expressed as a percentage) by Curray's
    [-@curray1956] formula tests the hypothesis of an unimodal
    distribution of orientations. If the $p$ value returned by the
    Rayleigh test is less than the threshold of 0.05, the null hypothesis
    of a random sampling from a uniform distribution is rejected. The
    formula for calculating the Vector Magnitude $L$ is as follows:

@@

Certain geological processes can have a different impact on objects depending on their
physical properties (*e.g.* silex versus bones in a water flow as in the
example files provided where bones have a strong preferential
orientation and are much more 'linear' in the Benn diagram [@thomas2019; @texier2020]). The natural processes models that you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find
the reference model data (with corresponding literature references)
which is used in the application in the 'Model reference data' subpanel.
In the rose diagram, axial data is considered (and not orientations) as
the Rayleigh test is perform on axial data [@curray1956] in the summary
table. You can download all the figures by selecting a file format
(.pdf, .jpeg and .png) and pressing the 'download' button (fig.
\ref{fig:figure_classical_method}E and F).

@@

-   use fabryka online: <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source code: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
-   Report bugs or any request: <https://github.com/marchaeologist/fabryka/issues> \smallbreak
-   github reprository doi of fabryka: <https://doi.org/10.5281/zenodo.15909666> \smallbreak
-   Zenodo repository of this article: <https://doi.org/10.5281/zenodo.16601423> \bigbreak

@@

2. Clone the fabryka repository from GitHub:
<https://github.com/marchaeologist/fabryka> or download the code in
"Code" and "Download ZIP" at the same address. Unzip the folder. This will give you the
latest version of the source code.

@@

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
Todisco on issues related to site formation processes. The reference
data used to build the models of the classical method were collected by
P. Bertran throughout his career. Pascal Bertran provided us with all these
data in order to feed the application. The development of the
application also benefited from the help of Rémi Lemoy, François Baleux and
Victor Baleux on mathematical issues. A first version of the application
was tested by Emmanuel Discamps and Aurélien Royer. This project is
supported by the "Service Valorisarion de la Recherche" of the
University Toulouse Jean Jaurès. Finally, the constructive comments of
Pascal Bertran, recommender of this article, and, of Frédéric
Santos, Nicolas Frerebeau and Alfonso Benito-Calvo, reviewers of this
article, considerably improved the manuscript and the web application. I
would like to thank all these reasearchers for their advice. \bigbreak
